NumLin: Linear Types for Linear Algebra

### 2019-05-26 write-up/paper/lang_and_examples.tex
Fix spacing
Fix punctuation
Incorporate final review feedback
Also reduce warnings.

\subsection{Syntax}

\begin{figure}[p]
    \begin{center}
    \[\def\arraystretch{1.3}
    \begin{array}{rcl}
        f & ::= & {'\!f\!c}
                \mid \Zf
                \mid f\ \Sf \\
                \\
        t & ::= & \Unit
                \mid \Bool
                \mid \Int
                \mid \Elt
                \mid f\ \Arr
                \mid f\ \Mat
                \mid \Bang t
                \mid {'\!f\!c}.\ t
                \mid t \otimes t'
                \mid t \multimap t' \\
                \\
        e & ::= & p \textrm{ (primitives)}
                \mid x \textrm{ (variable)}
                \mid \Let\ x = e\ \In\ e'
                \mid ()
                \mid \Let\ () = e\ \In\ e'
                \mid \ottkw{true}
                \mid \ottkw{false}
                \\ && \ottkw{if}\ e\ \ottkw{then}\ e_1\ \ottkw{else}\ e_2
                \mid k \textrm{ (integer)}
                \mid l \textrm{ (heap location)}
                \mid el \textrm{ (array element)}
                \\ && \Many\ v
                \mid \Let\ \Many\ x = e\ \In\ e'
                \mid \ottkw{fun}\ '\!f\!c \rightarrow e
                    \textrm{ (frac. perm. abstraction)}
                \\ && e\ [f] \textrm{ (frac. perm. specialisiation)}
                \mid (e, e')
                \mid \Let\ (a,b) = e\ \In\ e'
                \\ && \ottkw{fun}\ x : t \rightarrow e
                \mid e\ e'
                \mid \ottkw{fix} \: (g, x : t, e : t')
    \end{array} \]
    \end{center}
    \caption{Core fraction $f$, type $t$ and expression $e$ grammar of
        \lang. Values $v$ are a subset of the expressions, their full
        definition and a list of all primitives $p$ is in
        Appendix~\ref{sec:grammar_def}.}\label{fig:core_grammar}
\end{figure}

\begin{figure}[p]
\begin{center}
\[\def\arraystretch{1.3}
    \begin{array}{rcl}
    x[e] &
    \Rightarrow &
    \ottkw{get}\ \_\ x\ (e) \:\qquad \textrm{(similarly for matrices)}
\\
    x[e_1] := e_2 &
    \Rightarrow &
    \ottkw{set}\ x\ (e_1)\ (e_2) \quad \textrm{(similarly for matrices)}
\\
\\
    pat & ::= & () \mid x \mid {!x} \mid \Many\ pat \mid (pat, pat)
\\
    \Let\ !x = e_1\ \In\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{%
        \Let\ \Many\ x = e_1\ \In \\
        \Let\ \Many\ x = \Many\ (\Many\ x)\ \In\ e_2}
\\
    \Let\ \Many\ \langle pat_x \rangle\ = e_1\ \In\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{%
        \Let\ \Many\ x = x\ \In \\
        \Let\ \langle pat_x \rangle\ = x\ \In\ e_2}
\\
    \Let\ (\langle pat_a \rangle, \langle pat_b \rangle)\ = e_1\ \In\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{%
        \Let\ (a,b)\ = a\_b\ \In\
        \Let\ \langle pat_a \rangle\ = a\ \In \\
        \Let\ \langle pat_b \rangle\ = b\ \In\ e_2}
\\
    \ottkw{fun}\ (\langle pat_x \rangle : t) \rightarrow e &
    \Rightarrow &
    \ottkw{fun}\ (x : t) \rightarrow \Let\ \langle pat_x \rangle = x\ \In\ e
\\
\\
    arg & ::= & \langle pat \rangle : t\ \mid\ {'x} \textrm{ (fractional permission variable)}
\\
    \ottkw{fun}\ \langle arg_{1 .. n} \rangle \rightarrow e &
    \Rightarrow &
    \ottkw{fun}\ \langle arg_1 \rangle \rightarrow ..
    \ \ottkw{fun}\ \langle arg_n \rangle \rightarrow e
\\
    \Let\ f\ {\langle arg_{1 .. n} \rangle} = e_1\ \In\ e_2 &
    \Rightarrow &
    \Let\ f = \ottkw{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1\
    \In\ e_2
\\
    \Let\ !f\ {\langle arg_{1 .. n} \rangle} = e_1\ \In\ e_2 &
    \Rightarrow &
    \Let\ \Many\ f = \Many\ (\ottkw{fun}\ {\langle arg_{1 .. n} \rangle}
    \rightarrow e_1)\ \In\ e_2
\\
    \mathrm{fixpoint} & \equiv & \ottkw{fix}\ (f, x : t, \ottkw{fun}
    \ {\langle arg_{0 .. n} \rangle} \rightarrow e_1 : {t'} )
\\
    \Let\ \ottkw{rec}\ f\ (x : t)\ {\langle arg_{0 .. n} \rangle} : {t'} = e_1\ \In\ e_2 &
    \Rightarrow &
    \Let\ f = \mathrm{fixpoint}\ \In\ e_2
\\
    \Let\ \ottkw{rec}\ !f\ (x : t)\ {\langle arg_{0 .. n} \rangle} : {t'} = e_1\ \In\ e_2 &
    \Rightarrow &
    \Let\ \Many\ f = \Many\ \mathrm{fixpoint}\ \In\ e_2
    \end{array}
\]
\end{center}
\caption{Desugaring \lang.}\label{fig:lang_desugar}
\end{figure}

@@

The core type theory of \lang\ is a nearly off-the-shelf linear type
theory~\cite{barber1996}, supporting familiar features such as linear function
spaces $A \multimap B$ and tensor products $A \otimes B$. We adopt linearity --
the restriction that each program variable be used at most once -- since it
allows us to express purely functional APIs for numerical library routines that
mutate arrays and matrices~\cite{Wadler90}. Due to linearity, values cannot
alias and are only used once, which means that linearly-typed updates result in
no \emph{observable} mutation.

@@

\subsubsection{Intuitionism: ! and \highl{Many}}

@@

\subsubsection{Fractional Permissions}

@@

@@

Although a factorial function (Figure~\ref{fig:lang_factorial}) may seem like
an aggresively pedestrian first example, in a linearly typed language such as
\lang\ it represents the culmination of many features.

To simplifiy the design and implementation of \lang's type system, recursive
functions must have full type annotations (non-recursive functions need only
their argument types annotated). Its body is a closed expression (with respect
to the function's arguments), so it type-checks (since it does not capture any
linear values from its environment).

The only argument is \highl{!x : !int}. The ! annotation on \texttt{x} is a
syntactic convenience for declaring the value to used intuitionistically, its
full and precise meaning is described in Section~\ref{subsec:core_tt}.

@@

All numeric and boolean literals are implicitly wrapped in a \highl{Many} and
all primitives involving them return a \highl{!int}, \highl{!bool} or
\highl{!elt} (types of elements of arrays/matrices, typically 64-bit
floating-point numbers). The short-circuting \texttt{||} behaves in exactly the
same way as a boolean-valued \highl{if}-expression.

@@

An example of this consuming and re-binding is in \highl{let (row, !x1) =
row[i]}. Indexing is implemented as a primitive $\ottkw{get}:\ {'x}.\ {{'x}\
\Arr} \multimap {!\Int} \multimap {'x\ \Arr} \otimes {!\Elt}$.
Although fractional permissions can be passed around explicitly  (as done in
the recursive call), they can also be \emph{automatically inferred at call
sites}: \highl{row[i] == get _ row i} takes advantage of this convenience.

@@

Mutating an array is implemented similarly to indexing one -- a primitive
$\ottkw{set}:\ {\Zf\ \Arr} \multimap {!\Int} \multimap {!\Elt} \multimap
{\Zf\ \Arr}$. It consumes the original array and returns a new array with
the updated value.  \highl{let written = write[i] := <exp> } is just syntactic
sugar for \highl{let written = set write i <exp>}.

Since \highl{write: z arr} (where \Zf stands for $k=0$, representing a
fractional permission of $2^{-k} = 2^{-0} = 1$), we may mutate it, but since we
only need to read from \texttt{weights}, its fractional permission index can be
universally-quantified. In the recursive call, we see \texttt{\_} being used
explicitly to tell the compiler to \emph{infer} the correct fractional
permission based on the given arguments.

@@

We determine the types for these routines by consulting their documentation.
Each routine has a record of the expected aliasing behaviour and whether or not
it modifies or consumes its argument in any way. We use that to derive the
types in Figure~\ref{fig:primitive_types}. Since most of these low-level
routines are very careful not to do any allocation themselves, it is generally
very easy to give each a \lang\ type -- every argument that can modify/consume
its argument needs a full permission, and all others can be
fraction-polymorphic.  Taking Fortran as an example, it has a notion of
\texttt{in}, \texttt{out} and \texttt{inout} parameters. The latter two would
need full \Zf permissions; the first would be fraction-polymorphic.

@@

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../examples/square.lt}
    \caption{Linear regression (OLS): $\hat\beta =
        \mathbf{(X^T X)^{-1} X^T y}$}\label{fig:lang_square}
\end{figure}

@@

The line \highl{let answer <- new (m,n) [| x1 * x2 |]} is syntactic sugar for
first creating a new $m \times n$ matrix (\highl{let answer = matrix m n}) and
then storing the result of the multiplication in it (\highl{let ((x1, x2),
answer) = gemm 1. _ (x1, false) _ (x2, false) 0. answer}). \highl{false} means
the matrix should not be accessed with indices transposed.

@@

\begin{figure}[t]
\begin{center}
\[\def\arraystretch{1.3}
    \begin{array}{rcl}
    \Let\ v \leftarrow x[e]\ \In\ e &
    \Rightarrow &
    \Let\ (x, !v)\ = x[e]\ \In\ e \qquad \textrm{(similarly for matrices)}
\\
    \Let\ x_2 \leftarrow \ottkw{new}\ [|\ x_1\ |]\ \In\ e &
    \Rightarrow &
    \Let\ (x_1, x_2)\ = \ottkw{copyM}\ \_\ x_1\ \In\ e
\\
    \Let\ x_2 \leftarrow [|\ x_1\ |]\ \In\ e &
    \Rightarrow &
    \Let\ (x_1, x_2)\ = \ottkw{copyM\_to}\ \_\ x_1\ x_2\ \In\ e
    \end{array}
\]
\[
    M ::= X\ \mid\ X^T\ \mid\ \textrm{sym}( X )
\]
\end{center}
\begin{align*}
% new
    \Let\ &Y \leftarrow \ottkw{new}\ (n, k)\ [|\ \alpha M_1 M_2 \ |]\ \In\ e
    \Rightarrow \\
    & \Let\ Y\ = \ottkw{matrix}\ n\ k\ \In
    \ \Let\ Y \leftarrow [|\ \alpha M_1 M_2 + 0Y\ |]\ \In\ e
\\[0.1\baselineskip]
%syrk
    \Let\ &Y \leftarrow [|\ \alpha X X^T + \beta Y\ |]\ \In\ e
    \Rightarrow \\
    & \Let\ (X,Y)\ = \ottkw{syrk}\ \ottkw{false}
    \ \alpha\ \_\ X\ \beta\ Y\ \In\ e
\\[0.1\baselineskip]
    \Let\ &Y \leftarrow [|\ \alpha X^T X + \beta Y\ |]\ \In\ e
    \Rightarrow \\
    & \Let\ (X,Y)\ = \ottkw{syrk}\ \ottkw{true}
    \ \alpha\ \_\ X\ \beta\ Y\ \In\ e
%symm
\\[0.1\baselineskip]
    \Let\ &Y \leftarrow\ [|\ \alpha\,\textrm{sym}(X_1)\,X_2 + \beta Y\ |]\ \In\ e
    \Rightarrow \\
    & \Let\ ((X_1,X_2),Y)\ = \ottkw{symm}\ \ottkw{false}
        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \In\ e
\\[0.1\baselineskip]
    \Let\ &Y \leftarrow\ [|\ \alpha\,X_2\,\textrm{sym}(X_1) + \beta Y\ |]\ \In\ e
    \Rightarrow \\
    & \Let\ ((X_1,X_2),Y)\ = \ottkw{symm}\ \ottkw{true}
        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \In\ e
\\[0.1\baselineskip]
%gemm
    \Let\ &Y \leftarrow [|\ \alpha X_1^{T?} X_2^{T?} + \beta Y\ |]\ \In\ e
    \Rightarrow \\
    & \Let\ ((X_1,X_2),Y)\ =\ \ottkw{gemm}\ \alpha
        \ \_\ \left(X_1, \substack{\ottkw{true}\\\ottkw{false}}\right)
        \ \_\ \left(X_2, \substack{\ottkw{true}\\\ottkw{false}}\right)
        \ \beta\ Y\ \In\ e
\end{align*}

@@

However, the line \highl{let x_T_x <- new (m,m) [| x^T * x |]}, works for a
slightly different reason: that pattern is matched to a BLAS call to
(\highl{syrk true 1. x 0. x_T_x}), which only uses \texttt{x} once. Hence
\texttt{x} can appear \emph{twice} in the \emph{pattern} without any calls to
\texttt{share}.

@@

Figure~\ref{fig:lang_l1_norm_min} shows even more pattern-matching. Patterns of
the form \highl{let <id> <- [| beta * c + alpha * a * b |]} are also desugared
to \texttt{gemm} calls. Primitives like $\ottkw{transpose}:\ {'x}.\ {{'x}\ \Mat}
\multimap {{'x}\ \Mat} \otimes {\Zf\ \Mat}$ and $\ottkw{eye}:\ {!\Int}
\multimap {\Zf\ \Mat}$ allocate new matrices; \texttt{transpose} returns
the transpose of a given matrix and \texttt{eye k} evaluates to a $k \times k$
identity matrix.

We also see our first example of re-using memory for different matrices: like
with \texttt{to\_del} and \texttt{posv} in the previous example, we do not need
the value stored in \texttt{tmp\_5\_5} after the call to \texttt{gesv} (a
primitive similar to \texttt{posv} but for a non-symmetric first argument).
However, we can re-use its memory much later to store \texttt{answer} with
\highl{let answer <- [| 0. * tmp_5_5 + q_inv_u * inv_u_T |]}. Again, thanks to
linearity, the identifiers \texttt{q} and \texttt{tmp\_5\_5} are out of scope
by the time \texttt{answer} is bound. Although during execution, all three
refer to the same piece of memory, logically they represent different values
throughout the computation.

@@

    \item \highl{let new_r <- new [| r_2 |]} creates a copy of \texttt{r\_2}

@@

    \item a lot of memory re-use; the following sets of identifiers alias each other:
        \begin{itemize}
            \item \texttt{r\_1}, \texttt{r\_2} and \texttt{k\_by\_k}
            \item \texttt{data\_1} and \texttt{data\_2}
            \item \texttt{mu} and \texttt{new\_mu}
            \item \texttt{sigma\_hT} and \texttt{x}
        \end{itemize}

>>>>

@@

The core type theory of \lang\ is a nearly off-the-shelf linear type
theory~\cite{barber1996}, supporting familiar features such as linear function
spaces $A \multimap B$ and tensor products $A \otimes B$. We adopt linearity --
the restriction that each program variable be used exactly once -- since it
allows us to express purely functional APIs for numerical library routines that
mutate arrays and matrices~\cite{Wadler90}. Due to linearity, values cannot
alias and are only used once, which means that linearly-typed updates result in
no \emph{observable} mutation.

@@

\subsubsection{Intuitionism: ! and \texorpdfstring{\highl{Many}}{Many}}

@@

\subsubsection{Fractional Permissions}\label{subsubsec:frac_perm}

@@

\subsection{Syntax}\label{subsec:syntax}

\begin{figure}[p]
    \begin{center}
    \[\def\arraystretch{1.3}
    \begin{array}{rcl}
        f & ::= & {'\!f\!c}
                \mid \Zf
                \mid f\ \Sf \\
                \\
        t & ::= & \Unit
                \mid \Bool
                \mid \Int
                \mid \Elt
                \mid f\ \Arr
                \mid f\ \Mat
                \mid \Bang t
                \mid {'\!f\!c}.\ t
                \mid t \otimes t'
                \mid t \multimap t' \\
                \\
        e & ::= & p \textrm{ (primitives)}
                \mid x \textrm{ (variable)}
                \mid \Let\ x = e\ \In\ e'
                \mid ()
                \mid \Let\ () = e\ \In\ e'
                \mid \ottkw{true}
                \mid \ottkw{false}
                \\ && \ottkw{if}\ e\ \ottkw{then}\ e_1\ \ottkw{else}\ e_2
                \mid k \textrm{ (integer)}
                \mid l \textrm{ (heap location)}
                \mid el \textrm{ (array element)}
                \\ && \Many\ v
                \mid \Let\ \Many\ x = e\ \In\ e'
                \mid \ottkw{fun}\ '\!f\!c \rightarrow e
                    \textrm{ (frac. perm. abstraction)}
                \\ && e\ [f] \textrm{ (frac. perm. specialisation)}
                \mid (e, e')
                \mid \Let\ (a,b) = e\ \In\ e'
                \\ && \ottkw{fun}\ x : t \rightarrow e
                \mid e\ e'
                \mid \ottkw{fix} \: (g, x : t, e : t')
    \end{array} \]
    \end{center}
    \caption{Core fraction $f$, type $t$ and expression $e$ grammar of
        \lang. Values $v$ are a subset of the expressions, their full
        definition and a list of all primitives $p$ is in
        Appendix~\ref{sec:grammar_def}.}\label{fig:core_grammar}
\end{figure}

\begin{figure}[p]
\begin{center}
\[\def\arraystretch{1.3}
    \begin{array}{rcl}
    x[e] &
    \Rightarrow &
    \ottkw{get}\ \_\ x\ (e) \:\qquad \textrm{(similarly for matrices)}
\\
    x[e_1] := e_2 &
    \Rightarrow &
    \ottkw{set}\ x\ (e_1)\ (e_2) \quad \textrm{(similarly for matrices)}
\\
\\
    pat & ::= & () \mid x \mid {!x} \mid \Many\ pat \mid (pat, pat)
\\
    \Let\ !x = e_1\ \In\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{%
        \Let\ \Many\ x = e_1\ \In \\
        \Let\ \Many\ x = \Many\ (\Many\ x)\ \In\ e_2}
\\
    \Let\ \Many\ \langle pat_x \rangle\ = e_1\ \In\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{%
        \Let\ \Many\ x = x\ \In \\
        \Let\ \langle pat_x \rangle\ = x\ \In\ e_2}
\\
    \Let\ (\langle pat_a \rangle, \langle pat_b \rangle)\ = e_1\ \In\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{%
        \Let\ (a,b)\ = a\_b\ \In\
        \Let\ \langle pat_a \rangle\ = a\ \In \\
        \Let\ \langle pat_b \rangle\ = b\ \In\ e_2}
\\
    \ottkw{fun}\ (\langle pat_x \rangle : t) \rightarrow e &
    \Rightarrow &
    \ottkw{fun}\ (x : t) \rightarrow \Let\ \langle pat_x \rangle = x\ \In\ e
\\
\\
    arg & ::= & (\langle pat \rangle : t)\ \mid\ ({'x}) \textrm{ (fractional permission variable)}
\\
    \ottkw{fun}\ \langle arg_{1 .. n} \rangle \rightarrow e &
    \Rightarrow &
    \ottkw{fun}\ \langle arg_1 \rangle \rightarrow ..
    \ \ottkw{fun}\ \langle arg_n \rangle \rightarrow e
\\
    \Let\ f\ {\langle arg_{1 .. n} \rangle} = e_1\ \In\ e_2 &
    \Rightarrow &
    \Let\ f = \ottkw{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1\
    \In\ e_2
\\
    \Let\ !f\ {\langle arg_{1 .. n} \rangle} = e_1\ \In\ e_2 &
    \Rightarrow &
    \Let\ \Many\ f = \Many\ (\ottkw{fun}\ {\langle arg_{1 .. n} \rangle}
    \rightarrow e_1)\ \In\ e_2
\\
    \mathrm{fixpoint} & \equiv & \ottkw{fix}\ (f, x : t, \ottkw{fun}
    \ {\langle arg_{0 .. n} \rangle} \rightarrow e_1 : {t'} )
\\
    \Let\ \ottkw{rec}\ f\ (x : t)\ {\langle arg_{0 .. n} \rangle} : {t'} = e_1\ \In\ e_2 &
    \Rightarrow &
    \Let\ f = \mathrm{fixpoint}\ \In\ e_2
\\
    \Let\ \ottkw{rec}\ !f\ (x : t)\ {\langle arg_{0 .. n} \rangle} : {t'} = e_1\ \In\ e_2 &
    \Rightarrow &
    \Let\ \Many\ f = \Many\ \mathrm{fixpoint}\ \In\ e_2
    \end{array}
\]
\end{center}
\caption{Desugaring \lang.}\label{fig:lang_desugar}
\end{figure}

\begin{figure}[t]
\begin{center}
\[\def\arraystretch{1.3}
    \begin{array}{rcl}
    \Let\ v \leftarrow x[e]\ \In\ e' &
    \Rightarrow &
    \Let\ (x, !v)\ = x[e]\ \In\ e' \qquad \textrm{(similarly for matrices)}
\\
    \Let\ x_2 \leftarrow \ottkw{new}\ [|\ x_1\ |]\ \In\ e &
    \Rightarrow &
    \Let\ (x_1, x_2)\ = \ottkw{copyM}\ \_\ x_1\ \In\ e
\\
    \Let\ x_2 \leftarrow [|\ x_1\ |]\ \In\ e &
    \Rightarrow &
    \Let\ (x_1, x_2)\ = \ottkw{copyM\_to}\ \_\ x_1\ x_2\ \In\ e
    \end{array}
\]
\[
    M ::= X\ \mid\ X^T\ \mid\ \textrm{sym}( X )
\]
\end{center}
\begin{align*}
% new
    \Let\ &Y \leftarrow \ottkw{new}\ (n, k)\ [|\ \alpha M_1 M_2 \ |]\ \In\ e
    \Rightarrow \\
    & \Let\ Y\ = \ottkw{matrix}\ n\ k\ \In
    \ \Let\ Y \leftarrow [|\ \alpha M_1 M_2 + 0Y\ |]\ \In\ e
\\[0.1\baselineskip]
%syrk
    \Let\ &Y \leftarrow [|\ \alpha X X^T + \beta Y\ |]\ \In\ e
    \Rightarrow \\
    & \Let\ (X,Y)\ = \ottkw{syrk}\ \ottkw{false}
    \ \alpha\ \_\ X\ \beta\ Y\ \In\ e
\\[0.1\baselineskip]
    \Let\ &Y \leftarrow [|\ \alpha X^T X + \beta Y\ |]\ \In\ e
    \Rightarrow \\
    & \Let\ (X,Y)\ = \ottkw{syrk}\ \ottkw{true}
    \ \alpha\ \_\ X\ \beta\ Y\ \In\ e
%symm
\\[0.1\baselineskip]
    \Let\ &Y \leftarrow\ [|\ \alpha\,\textrm{sym}(X_1)\,X_2 + \beta Y\ |]\ \In\ e
    \Rightarrow \\
    & \Let\ ((X_1,X_2),Y)\ = \ottkw{symm}\ \ottkw{false}
        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \In\ e
\\[0.1\baselineskip]
    \Let\ &Y \leftarrow\ [|\ \alpha\,X_2\,\textrm{sym}(X_1) + \beta Y\ |]\ \In\ e
    \Rightarrow \\
    & \Let\ ((X_1,X_2),Y)\ = \ottkw{symm}\ \ottkw{true}
        \ \alpha\ \_ \ X_1 \_\ X_2\ \beta\ Y\ \In\ e
\\[0.1\baselineskip]
%gemm
    \Let\ &Y \leftarrow [|\ \alpha X_1^{T?} X_2^{T?} + \beta Y\ |]\ \In\ e
    \Rightarrow \\
    & \Let\ ((X_1,X_2),Y)\ =\ \ottkw{gemm}\ \alpha
        \ \_\ \left(X_1, \substack{\ottkw{true}\\\ottkw{false}}\right)
        \ \_\ \left(X_2, \substack{\ottkw{true}\\\ottkw{false}}\right)
        \ \beta\ Y\ \In\ e
\end{align*}

As described in Section~\ref{subsubsec:frac_perm}, fractional permissions $f$
are either variables $'\!f\!c$, $\Zf$ ($2^0$) or $f\ \Sf$ ($2^{-(f + 1)}$).

Types are either simple ($\Unit$, $\Bool$, $\Int$, $\Elt$), indexed
by fractional permission ($f\ \Arr$, $f\ \Mat$) or compound (!
constructor for intuitionism, universally-quantifying over a
fractional-permission $'\!f\!c$ in $t$ for fraction-polymorphic types, pairs,
linear functions).

Expressions are standard with the exception of $'\!f\!c.\ t$ introduction
(fractional permission abstraction) and elimination (fractional permission
specialisation) forms.

The ! annotation on variables is a syntactic convenience for declaring the
variable to used intuitionistically; why it desugars the way it does is
explained in Section~\ref{subsec:core_tt}.

Array/matrix indexing and duplication (non-destructive and destructive) also
have special syntax to lessen the syntactic overhead of re-binding identifiers.
Furthermore, to aid readability, there is support for using BLAS methods via
conventional-looking matrix expressions.

In particular, the syntax \highl{let y <- new (m,n) [| alpha * x1 * x2 |]} is
syntactic sugar for first creating a new $m \times n$ matrix (\highl{let y =
matrix m n}) and then storing the result of the multiplication in it
(\highl{let ((x1, x2), y) = .. in ..}).

Note that, the pattern \highl{let y <- [| x^T * x + beta * y |]} translates to
(\highl{syrk true 1. x beta y}), which uses \texttt{x} once only.

@@

Although a factorial function (Figure~\ref{fig:lang_factorial}) may seem like
an aggressively pedestrian first example, in a linearly typed language such as
\lang\ it represents the culmination of many features.

To simplify the design and implementation of \lang's type system, recursive
functions must have full type annotations (non-recursive functions need only
their argument types annotated). The body of the factorial function is a closed
expression (with respect to the function's arguments), so it type-checks (since
it does not capture any linear values from its environment).

The only argument is \highl{!x : !int}. As explained before
(Section~\ref{subsec:syntax}), this declares \texttt{x} to be used
intuitionistically.

@@

All numeric and boolean literals are implicitly wrapped in a \highl{Many} and
all primitives involving them return a \highl{!int}, \highl{!bool} or
\highl{!elt} (types of elements of arrays/matrices, typically 64-bit
floating-point numbers). The short-circuiting \texttt{||} behaves in exactly the
same way as a boolean-valued \highl{if}-expression.

@@

An example of this consuming and re-binding is in \highl{let (row, !x1) =
row[i]}. Indexing is implemented as a primitive $\ottkw{get}:\ {'x}.\ {{'x}\
\Arr} \multimap {!\Int} \multimap {'x\ \Arr} \otimes {!\Elt}$.
Although fractional permissions can be passed around explicitly  (as done in
the recursive call), they can also be \emph{automatically inferred at call
sites}: \highl{row[i]} $\Rightarrow$ \highl{get _ row i} takes advantage of
this convenience.

@@

Mutating an array is implemented similarly to indexing one -- a primitive
$\ottkw{set}:\ {\Zf\ \Arr} \multimap {!\Int} \multimap {!\Elt} \multimap
{\Zf\ \Arr}$. It consumes the original array and returns a new array with
the updated value.

Since \highl{write: z arr} (where \Zf\ stands for $k=0$, representing a
fractional permission of $2^{-k} = 2^{-0} = 1$), we may mutate it, but since we
only need to read from \texttt{weights}, its fractional permission index can be
universally-quantified. In the recursive call, we see \texttt{\_} being used
explicitly to tell the compiler to \emph{infer} the correct fractional
permission based on the given arguments.

@@

We determine the types for these routines by consulting their documentation.
Each routine has a record of the expected aliasing behaviour and whether or not
it modifies or consumes its argument in any way. We use that to derive the
types in Figure~\ref{fig:primitive_types}. Since most of these low-level
routines are very careful not to do any allocation themselves, it is generally
very easy to give each a \lang\ type -- every argument that can modify/consume
its argument needs a full permission, and all others can be
fraction-polymorphic.  Taking Fortran as an example, it has a notion of
\texttt{in}, \texttt{out} and \texttt{inout} parameters. The latter two would
need full \Zf\ permissions; the first would be fraction-polymorphic.

@@

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../examples/square.lt}
    \caption{Squaring a matrix}\label{fig:lang_square}
\end{figure}

@@

@@

@@

Note that \texttt{x} can appear \emph{twice} in the \emph{pattern} without any
calls to \texttt{share} because the pattern is matched to a BLAS call to
\highl{syrk true 1. x 0. x_T_x}, which uses \texttt{x} once only.

@@

Primitives like $\ottkw{transpose}:\ {'x}.\ {{'x}\ \Mat}
\multimap {{'x}\ \Mat} \otimes {\Zf\ \Mat}$ and $\ottkw{eye}:\ {!\Int}
\multimap {\Zf\ \Mat}$ allocate new matrices; \texttt{transpose} returns
the transpose of a given matrix and \texttt{eye k} evaluates to a $k \times k$
identity matrix.

We also see our first example of re-using memory for different matrices: like
with \texttt{to\_del} and \texttt{posv} in the previous example, we do not need
the value stored in \texttt{tmp\_n\_n} after the call to \texttt{gesv} (a
primitive similar to \texttt{posv} but for a non-symmetric first argument).
However, we can re-use its memory much later to store \texttt{answer} with
\highl{let answer <- [| 0. * tmp_n_n + q_inv_u * inv_u_T |]}. Again, thanks to
linearity, the identifiers \texttt{q} and \texttt{tmp\_n\_n} are out of scope
by the time \texttt{answer} is bound. Although during execution, all three
refer to the same piece of memory, logically they represent different values
throughout the computation.

@@

@@

    \item a lot of memory re-use; the following sets of identifiers alias each other:
        \begin{itemize}
            \item \texttt{r\_1}, \texttt{r\_2} and \texttt{k\_by\_k}
            \item \texttt{data\_1} and \texttt{data\_2}
            \item \texttt{mu} and \texttt{new\_mu}
            \item \texttt{sigma\_hT} and \texttt{x}.
        \end{itemize}

### 2019-05-26 write-up/paper/paper.tex
Fix formatting
Make stuff consistent
Syntax & spacing, with a few small edits from Reviewer C's review.
Incorporate final review feedback
Also reduce warnings.

\supplement{\url{www.github.com/dc-mak/lt4la}}

%\nolinenumbers %uncomment to disable line numbering

@@

%Editor-only macros:: begin (do not touch as author)%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}

@@

\newcommand{\Unit}{\ottkw{unit}}%
\newcommand{\Bang}{\ottkw{!}}
\newcommand{\Bool}{\ottkw{bool}}%
\newcommand{\Int}{\ottkw{int}}%
\newcommand{\Elt}{\ottkw{elt}}%
\newcommand{\Mat}{\ottkw{mat}}%
\newcommand{\Zf}{\ottkw{Z}}%
\newcommand{\Sf}{\ottkw{S}}%
\newcommand{\Many}{\ottkw{Many}}%
\newcommand{\dom}{\mathrm{dom}}%
\newcommand{\empH}{\emptyset}%

>>>>

\supplement{\url{www.github.com/dc-mak/NumLin}}

\acknowledgements{We would like to thank Stephen Dolan for his advice and
support with the implementation and evaluation of NumLin. We would also like to
thank the (anonymous) reviewers for their feedback and suggestions.}

\nolinenumbers %uncomment to disable line numbering

@@

%Editor-only macros:: begin (do not touch as author)%
\EventEditors{Alastair F. Donaldson}
\EventNoEds{1}
\EventLongTitle{33rd European Conference on Object-Oriented Programming (ECOOP 2019)}
\EventShortTitle{ECOOP 2019}
\EventAcronym{ECOOP}
\EventYear{2019}
\EventDate{July 15--19, 2019}
\EventLocation{London, United Kingdom}
\EventLogo{}
\SeriesVolume{134}
\ArticleNo{9}

@@

\newcommand{\Unit}{\ottkw{unit}}%
\newcommand{\Bang}{\ottkw{!}}
\newcommand{\Bool}{\ottkw{bool}}%
\newcommand{\Int}{\ottkw{int}}%
\newcommand{\Elt}{\ottkw{elt}}%
\newcommand{\Arr}{\ottkw{arr}}%
\newcommand{\Mat}{\ottkw{mat}}%
\newcommand{\Zf}{\texttt{z}}%
\newcommand{\Sf}{\texttt{s}}%
\newcommand{\Let}{\ottkw{let}}%
\newcommand{\In}{\ottkw{in}}%
\newcommand{\Many}{\ottkw{Many}}%
\newcommand{\dom}{\mathrm{dom}}%
\newcommand{\empH}{\emptyset}%
